The role of aneuploidy in the evolution of cancer drug resistance

### 2024-01-20 EvolutionaryRescue1.tex
Add files via upload

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then exposed to chemotherapeutic agents such as vemurafenib. When compared to wildtype cancer cell the reversine treated cells had higher proliferation in the environment altered by anti-cancer drugs. Similarly, \citet{lukow2021chromosomal} observed that cancer cells which have been induced to be aneuploid have an advantage when compare to wildtype cell even while they had lower fitness before the onset of chemotherapy. A proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is bu antagonizing cell division  which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}

 An important aspect of aneuploidy is the rate with which cell acquire it, which is several orders of magnitude higher when compare to mutation rate \citep{bakker2023predicting}. Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of missegration in order to combat cancer \citep{lee2016effects}.

@@

Here we build on this previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. Besides the probability of evolutionary rescue we calculate the calculate the mean time that an initially declining tumor cell population reaches the initial population size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$. We assume that both chromosomal missegregation and mutation occur during the process of mitosis. Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$. We assume that wildtype cells acquire aneuploidy before the onset of therapy with rate $\tilde{u}\lambda_w$ and that aneuploid cells have a fitness cost $s$ in the drug-free environment.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Stochastic simulations} 
Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.

@@

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log10}{\log\left(\lambda_m / \mu_m\right)}\right\rfloor + 1.
\end{equation*}

@@

@@

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improved the fitness of cancer cells in an environment altered by chemotherapeutic drugs. We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. \citet{ippolito2021gene} observed that trisomy 2 and 6 are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a cancer is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more then one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the percentage of cancer cells with favorable aneuploidy as $2/22\times0.15\approx 0.0136$ (i.e. $1.36\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy) which we select as the value for $f$ (this gives us a fitness cost of $s= 7.3\times10^{-3}\text{ days}^{-1}$). We note that when we refer to wildtype cancer cells we include those cells which have any aneuploidy except trisomy 2 and 6 as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by therapeutics such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources. This is a good approximation as we assume that the drug cause the wildtype population to decline rapidly and, as a result, density effect can be neglected.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= \lambda_a,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K},\\
\lambda_m' &= \lambda_m,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K}.
\end{align*}
where $K$ is the maximum carrying capacity. The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$.

@@

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
\begin{equation} \label{eq:rescue_prob} 
\presc = 
1-\left(1-p_w\right)^N \approx
1-\e^{-Np_w} = 
1-e^{-N/N^*} .
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size. 
\\
In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (partially tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob,rescue_threshold}). Increasing the the missegregation rate $u$ and aneuploid growth rate $\Delta_a$ reduces the threshold tumor size and facilitates the evolution of drug resistance (\Cref{rescue_threshold}).

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$. We observe that aneuploidy has a significant contribution towards the tumor overcoming chemotherapeutic drugs. 
% TODO PLUG IN REAL VALUES ok

@@

Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction-- (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time). However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time. Of particular interest is the distribution of the evolutionary rescue time and recurrence time in the parameter regime  $\Delta_a T^*\ll1$ (partially tolerant aneuploids) for which we select  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.

\paragraph{Recurrence time.}

In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time. \Cref{proliferationFigure} shows the agreement between our approximations and simulations. For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean proliferation time becomes constant.

\paragraph{Distribution of evolutionary rescue time -- with and without aneuploidy.}

>>>>

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to chemotherapeutic agents such as vemurafenib. When compared to wildtype cancer cell the reversine treated cells had higher proliferation in the environment altered by anti-cancer drugs. Similarly, \citet{lukow2021chromosomal} observed that cancer cells which have been induced to be aneuploid have an advantage when compare to wildtype even while they had lower fitness before the onset of chemotherapy. A proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division  which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}

 An important aspect of aneuploidy is the rate with which cells acquire it, which is several orders of magnitude higher when compare to the mutation rate \citep{bakker2023predicting}. Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of missegration in order to combat cancer \citep{lee2016effects}.

@@

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$. We assume that both chromosomal missegregation and mutation occur during the process of mitosis. Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \, (A), \Delta_a\ll0\, (B),\Delta_a\approx0\, (C),\Delta_a\gg0\, (D)$. We assume that wildtype cells acquire aneuploidy before the onset of therapy with rate $\tilde{u}\lambda_w$ and that aneuploid cells have a fitness cost $s$ in the drug-free environment.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Stochastic simulations} 
Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.

@@

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log10}{\log\left(\lambda_m / \mu_m\right)}\right\rfloor + 1.
\end{equation*}
where we obtain the threshold mutant size by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$, where $p_m=\Delta_m/\lambda_m$ is the probability that a single mutant escapes stochastic extinction  (see Apendix A).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Bootstrapping}
We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtain from simulations. We then compute the interval which contain $95\%$ of the means of each resampled set.

@@

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improved the fitness of cancer cells in an environment altered by chemotherapeutic drugs. We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. \citet{ippolito2021gene} observed that trisomy 2 and 6 are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more then one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the percentage of cancer cells with favorable aneuploidy as $2/22\times0.15\approx 0.0136$ (i.e. $1.36\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy) which we select as the value for $f$ (this gives us a fitness cost of $s= 7.3\times10^{-3}\text{ days}^{-1}$). We note that when we refer to wildtype cancer cells we include those cells which have any aneuploidy except trisomy 2 and 6 as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by therapeutics such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources. This is a good approximation as we assume that the drug cause the wildtype population to decline rapidly and, as a result, density effect can be neglected.
A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= \lambda_a,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K},\\
\lambda_m' &= \lambda_m,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K}.
\end{align*}
where $K$ is the maximum carrying capacity. The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
\begin{equation} \label{eq:rescue_prob} 
\presc = 
1-\left(1-p_w\right)^N \approx
1-\e^{-Np_w} = 
1-e^{-N/N^*} .
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size. 
\\
In Appendix A, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that 
\begin{equation} \label{eq:N_m}
N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w} \cdot \frac{\lambda_m}{\Delta_m} .
\end{equation}
Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob,rescue_threshold}). Increasing the the missegregation rate $u$ and aneuploid growth rate $\Delta_a$ reduces the threshold tumor size and facilitates the evolution of drug resistance (\Cref{rescue_threshold}).

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case. In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation. This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$. We note that for this parameter regime we have $N_a^*\approx4.4\times10^6$ and $N_m^*\approx4\times10^7$ and this allows us to define the following categories: small tumors have size $N<N_a^*$, intermediate sized tumors have $N_a^*<N<N_m^*$ and large tumors have size $N>N_m^*$ . For the remaining of this paper we use these definitions when referring to small, intermediate or large tumors. Interestingly, the threshold between small and intermediate tumors is similar to the detection threshold of $4.19\times10^6$ for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time). However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time. Of particular interest is the distribution of the evolutionary rescue time and recurrence time in the parameter regime  $\Delta_a T^*\ll1$ (partially tolerant aneuploids) for which we select  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot,EvolutionaryRescueTimeComplete} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes. Additionally, we note that $\tau_a$ has the following asymptotic behaviour:
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \left(v\lambda_wNp_m\right)^{-1}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
Interestingly, for small tumors the mean evolutionary rescue time is a function of the wildtype and aneuploid growth rates and independent of all of other model parameters including tumor size $N$ (\Cref{EvolutionaryRescueTimeComplete}). Increasing either the wildtype or aneuploid growth rates lead to an increase in the mean time. This can be interpreted that by increasing either growth rates the wildtype or aneuploid cells will last longer and will be able to produce a rescue mutation at latter times. Using the parameters that we have selected we observe that $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$ such that the mean evolutionary rescue time is determined by the aneuploid growth rate, i.e. $\tau_a\approx1/\abs{\Delta_a}$. For large tumors, we observe that the evolutionary rescue time is exponentially distributed with rate $v\lambda_wNp_m$ which is independent of parameters characterizing aneuploids or their production ($u$, $\lambda_a$, and $\Delta_a$). Increasing the mutation rate per mitosis $v\lambda_w$ leads to a faster appearance of a rescue mutation. Note that in the second case increasing the tumor size leads to smaller mean evolutionary rescue time as the wildtype is able to produce a rescue lineage faster.

When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \Cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
\paragraph{Recurrence time.}
